---
title: 'Mutations in SARS-CoV-2 Spike Protein and its Role in Transmissibility'
author: "Lewis Hendrianto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
csl: bioinformatics.csl
bibliography: references.bib
params:
  gff_file_path: "data/metadata/sars_refgenome_annotation.gff"
  vcf_dir_path: "data/11_vcf_output_for_R"
  sra_runtable_path: "data/00_sra_runtable/filtered_sra_table.txt"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE)
```

# Background and Overview

The World Health Organization (WHO) declared coronavirus 19 (COVID-19), a pandemic on March 11, 2020. (https://www.yalemedicine.org/news/covid-timeline) Severe acute respiratory syndrome coronavirus 2, also known as SARS-CoV-2, is a virus from the coronavirus family. SARS-CoV-2 causes this disease. COVID-19 is a respiratory disease primarily transmitted through aerosols. Preventative and therapeutic approaches are in development to combat this virus, and an mRNA vaccine has become widely available for the public. (https://www.frontiersin.org/articles/10.3389/fcimb.2020.587269/full)
Before understanding the function of the virus, it is essential to understand the structure of a virus. In general, each virus contains a viral genome that resides in a shell called the capsid. Additionally, the coronavirus contains a lipid membrane envelope that surrounds the capsid. Spike glycoproteins surround the lipid envelope, functioning as a gateway to enter the target host. The spike proteins bind to a specific receptor to the host, for SARS-CoV-2, its target is the angiotensin-converting enzyme 2 (ACE2). Upon binding, the virus will permeate the host cell’s membrane, inserting the viral genome. The viral genome gets replicated and expressed, producing viral proteins infect other cells ultimately causing COVID-19. (https://www.khanacademy.org/science/biology/biology-of-viruses/virus-biology/a/intro-to-viruses) Moreover, the spike proteins could undergo mutations, affecting is permeability and transmissibility towards the target host. Consequently, the SARS-CoV-2 has faced several mutations, acquiring SNPs in multiple positions. With the data collected through BioProject PRJNA793894, it has been identified that the S:665Y substitution causes these effects. This report will take the data and examine how the SNPs affected the viral entry and replication of the newly obtained variants. On November 8, 2021, due to SNPs, the first confirmed sample of the Omicron variant is collected in South Africa. In less than a month, the variant of concern (VOC) made its way across the world to the United States on December 1st 2021. Rapidly, the Omicron variant became one of the most dominant strains in the world because of the SNP in the spike protein. As of November 2022, Omicron is responsible for over 97% of COVID-19 transmissions according to the Centers for Disease Control and Prevention. As a result, this VOC takes in account for the majority of covid cases and deaths since its emergence. (https://health.ucdavis.edu/coronavirus/covid-19-information/omicron-variant)(https://covid.cdc.gov/covid-data-tracker/#variant-proportions)
The primary goal of this analysis were to identify the position of each SNP and interpret how they enhanced the transmissibility of this VOC compared to the original variant. Although the data collected from the biosample was from June 2021 through November 2021, the report discovered how the SNP has impacted the United States in recent times. The approach of the report was accomplished through a combination of bash scripts and R code. The provided code took the downloaded BioProject’s SRA runtable metadata and executed several ramifications. The 89 Illumina samples in the runtable was deconstructed, trimmed, and polished to produce VCF files that exposed the SNPs, quality scores, and other interpretable data that allowed a close analysis. Finally, a Makefile compiled the analysis and produced the report as a PDF.

# Methods

# Results

The BioProject collected 89 different samples, from four different isolates at the Icahn School of Medicine at Mount Sinai, located in New York, United States. The isolates originated from a nasal wash or nasal turbinate from Mesocricetus auratus and viral supernantants with pneumocyte infected cells and without. (Table 1). The samples were collected in the months of June 2021 and November 2021. Although a number of the collection dates were before the identification of Omicron, the same SNPs were present in the collection dates during Omicron’s emergence. Shortly after, the BioProject was released on January 3rd, 2021, past Omicron’s first invasion into the United States. Regarding the spike protein gene S, the SNP from nucleotide C to T remained conserved in both month’s of the collection date, occurring at the same chromosomal position 23525. Correspondingly, the SNPs present in noncoding regions and the SNP for the ORF8 gene also remained conserved. (Table 3). Parsing through the SRA runtable with the bash scripts, it was determined that the vast majority of SNPs in the sample were reliable and certain. The vast majority of the samples possessed SNPs that were confirmed. The samples almost consistently had a score greater than 220, with the some scoring in the range of 190 to 220. (Figure 1, Figure 2). For the latter, they were ruled out of the analysis due to the poor quality scores of 134 or lower. The poor SNPs mainly occurred at position 23525, with one changing to base pair A at 18883, and a change to C at 29737. (Table 4) It has been established that there are four certain SNPs that have occurred in SARS-CoV-2. The most common SNP changes the nucleotide C to T, and the other changing from T to C.However, only one SNP affects the spike protein’s gene, at 23525. As for the other affected gene, the 28144 SNP of base pair C is shown to alter the ORF8 gene. SNPs at position 8782 and 18060 contains the same SNP as the spike protein SNP, but it is said to not affect any particular gene. (Table 2). It has been identified that the SNPs are found in the Omicron variant. As a result, the deaths starting from December 1st, 2021, the day Omicron was identified in the united states, were recorded until November 23, 2022. In the time span, there has been 292, 973 confirmed deaths in the United States due to COVID-19. (Figure 3). Although the deaths do not explicity identify the variant to be the cause of death, we should consider that over 97% confirmed cases of COVID-19 are from Omicron since its emergence.

# Discussion

```{r load-packages-and-functions}
library("vcfR")
library("ggplot2")
library("dplyr")
library("ggthemes")
library("tidyr")
library("readr")

# this loads all function files in the code/functions directory
function_paths <- list.files("code/functions", full.names = TRUE)
purrr::walk(function_paths, source)
```

```{r load-vcf-data}
# load in, tidy, and stack all of the VCF files in the target directory
# these are the output of the bash pipeline
stacked_vcfs <- parse_tidy_and_stack_vcfs(
  vcf_dir_path = params$vcf_dir_path)

# load in the gff file with genome annotations for SARS-CoV-2
gff <- read_gff(gff_file_path = params$gff_file_path)

# pull out just the named genes from the annotation file
gene_table <- extract_genes_from_gff(annotation_object = gff)

# combine the stacked vcf table with the gene names and the metadata from
# the sra runtable file
vcf_with_metadata <- add_genes_metadata_to_vcfstack(
  sra_runtable_path = params$sra_runtable_path,
  stacked_vcf = stacked_vcfs,
  cleaned_genes_table = gene_table)
vcf_with_metadata$qual <- as.integer(vcf_with_metadata$qual)
colnames(vcf_with_metadata) <- gsub(" ", "_", colnames(vcf_with_metadata))
covid_tabular_data <- read_csv(file = "data/raw_data/covid_tabular_data.csv")

```

# Figures

```{r dot-plot}
# create a dot plot mapping out all of the samples collection date and
# their quality score
dot_plot <- ggplot(vcf_with_metadata[which(vcf_with_metadata$qual > 150),],
                       aes(x = Collection_Date,
                           y = qual)) +
  geom_point(col = "black", size = 1) +
  scale_x_date(date_labels = "%Y-%b-%d",date_breaks  = "1 month") +
  labs(x = "Date of sample collection",
       y = "Quality score of sample",
       title = "The collection date and quality score of each sample") +
    theme(axis.text.x = element_text(angle = 0, size = 7)) +
  theme(axis.text.y = element_text(angle = 0, size = 7)) +
  theme(plot.title = element_text(hjust = 0.5))

dot_plot
```

**Figure 1** The collection date of the samples and their quality scores

```{r SNP-quality-plot}
# create a flipped bar plot showing which SNPs are quality SNPs and which
# should be excluded
bar_plot <- ggplot(data = vcf_with_metadata,
                    aes(x = alt,
                        y = qual)) +
  geom_col(position = "dodge") +
  labs(x = "DNA base",
       y = "Quality score",
       title = "Quality of SNP") +
    theme(axis.text.x = element_text(angle = 0, size = 7)) +
  theme(axis.text.y = element_text(angle = 0, size = 7)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_discrete(guide = guide_axis(n.dodge = 1.75)) +
  coord_flip()

bar_plot
```

**Figure 2** The quality and confidence of the SNPs in the BioProject

```{r time-line plot}
# create a time-line plot that shows the deaths over nov 2021 until now
time_plot <- ggplot(data = covid_tabular_data,
                    aes(x = Date)) +
  geom_line(aes(y = Total_Deaths)) +
  geom_point(aes(y = Total_Deaths)) +
  labs(x = "Date",
       y = "Number of Deaths") +
  scale_x_date(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 60, size = 5, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)

time_plot
```

**Figure 3** The cumulative deaths since the emergence of Omicron

# Tables

```{r sample-bases-table}
# create a table showing the samples and what the isolation source is, and how
# many bases they have

vcf_with_metadata %>%
  select(sample, Isolation_source, geo_loc_name) %>% unique() %>%
  knitr::kable(col.names = c("Sample",
                             "Isolation source",
                             "Number of Bases"), row.names = FALSE,
               align = "lll")

```

**Table 1** The test samples and where they originated from

```{r snp-table}
# create a table mapping out where the SNP occurs

vcf_with_metadata %>%
  select(pos, ref, alt, gene) %>% unique() %>%
  knitr::kable(col.names = c("Chromosome position",
                             "Ref. Nucleotide",
                             "SNP",
                             "Gene affected"),
               row.names = FALSE,
               align = "lccc",
               na.rm = TRUE)

```

**Table 2**: Where each SNP occurs

```{r date-table}
# create a table showing when the samples were collected and released

vcf_with_metadata %>%
  arrange(vcf_with_metadata$Collection_Date) %>%
  select(Collection_Date, ReleaseDate, alt, gene, pos) %>% unique() %>%
  knitr::kable(col.names = c("Collection Date",
                             "Release Date",
                             "SNP",
                             "Gene",
                             "Position"),
               row.names = FALSE, align = "ll")

```

**Table 3**: The samples collection date and release date

```{r bad-SNP-table}
vcf_with_metadata[which(vcf_with_metadata$qual < 150),] %>%
  arrange((vcf_with_metadata[which(vcf_with_metadata$qual < 150),])$pos) %>%
  select(pos, alt, qual) %>% 
  unique() %>%
  knitr::kable(col.names = c("Chromosome position",
                             "SNP",
                             "Quality score"), row.names = FALSE, align = "lcc")
```

**Table 4**: The quality score of the SNPs that were excluded in the analysis

# Sources Cited